---
title: "hte"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hte}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(pmsims)
library(predieval)
library(benchtm)
```


## Motivation
Heterogeneous treatment effects occur when people with different characteristics get more or less benefit from interventions. Modelling heterogeneous treatment effects can facilitate personalised medicine, allowing the targeting of interventions to those who would benefit most.

Sample size is important to ensure sufficient power to detect hetrogeneous treatment effects and ensure that the models that predict hetrogenous treatment effects are not overfit to the data. 


### Defining the data generator
For this example we will use the **benchtm** R-package. This package was created to allow the simualtion of data from randomised trials when hetrogeneous treatment effects are present. 

I believe the prop = "X2" option when generating the treatment variable trt introduces confounding by X2.

```{r}
my_data_generator <- function(n) {
  
  # 1. Generate X distribution
  x_data <- tryCatch({
    benchtm::generate_X_dist(n = n, p = 10, rho = 0.5)
  }, error = function(e) {
    stop(sprintf("Error in generate_X_dist at n = %d: %s", n, e$message), call. = FALSE)
  })
  x_data <- x_data[, names(x_data) != "X6"] # removing X6 as catagorical variable simulated with varying numbers of levels which causes errors when differences occurs between test and train.
  
  # Process X data
  x_data_numeric <- model.matrix(~., data = x_data)[,-1] |> data.frame()
  
  # 2. Generate Treatment
  trt <- tryCatch({
    benchtm::generate_trt(n = nrow(x_data_numeric), type = "random", X = x_data_numeric, prop = "X2")
  }, error = function(e) {
    stop(sprintf("Error in generate_trt at n = %d: %s", n, e$message), call. = FALSE)
  })
  
  # 3. Generate Outcome (Y)
  y_data <- tryCatch({
    benchtm::generate_y(
      x_data_numeric, 
      trt = trt,
      prog = "0.5*((X1Y=='Y')+X3)",
      pred = "X3>0", # this defines how treatment effect heterogeneity is introduced.
      b0 = 0, 
      b1 = 1,
      type = "continuous", 
      sigma_error = 3
    )
  }, error = function(e) {
    stop(sprintf("Error in generate_y at n = %d: %s", n, e$message), call. = FALSE)
  })
  
  return(y_data)
}

example_data <- my_data_generator(n = 10)
print(example_data)
```

### Defining the model function
Next we need to define the model function. Our model will be a linear model with all the covariates included as main terms, as well as their pairwise interactions.

```{r}
my_model_function = function(data) {
  x_names <- names(data)[grepl("^X", names(data))]
  main_terms <- paste(x_names, collapse = " + ") 
  interaction_terms <- paste("trt*", x_names, sep = "", collapse = " + ") 
      formula <- glue::glue("Y ~ trt + {main_terms} + {interaction_terms}")
      my_model = lm(data = data, formula = as.formula(formula))
      return(my_model)
}
      
example_data <- my_data_generator(n = 100)
example_fitted_model <- my_model_function(example_data)
print(example_fitted_model)
```


### The metric function

For the perofrmance metric we carry out a joint test of all the interaction terms being zero (ie. no hetrogenous treatment effects) and return 1 if the test is statistically significant at the 5% level and 0 otehrwise.


```{r}
metric_test_for_hte <- function(test_data, fitted_model, model_name) {
  
  model_data <- model.frame(fitted_model)
  x_names <- names(model_data)[grepl("^X", names(model_data))]
  
  # Construct reduced model (no interactions)
  main_terms <- paste(x_names, collapse = " + ")
  reduced_formula <- paste0("Y ~ trt + ", main_terms)
  
  reduced_model <- lm(
    data = model_data,
    formula = as.formula(reduced_formula)
  )
  
  # Likelihood ratio test via ANOVA (nested models)
  lrt <- anova(reduced_model, fitted_model)
  
  p_val <- lrt$`Pr(>F)`[2]
  ifelse(p_val < 0.05, 1, 0)
}

```

### Sense check: Large sample performance
We would expect statistical power to be 1 when the sample size is very large. Model perofmrance would be 1.

```{r}

  set.seed(1234)
  large_sample_data <- my_data_generator(n = 100000)
  test_data <- my_data_generator(n = 10) # unused so set to be very small
  test_model <- my_model_function(large_sample_data)
  large_sample_performance <- metric_test_for_hte(test_data, test_model, "multivariable regression")
  print(large_sample_performance)
```

### Running Simulate Custom
We now run simulate custom. As we are looking to calculate statistical power we use the mean criterion and set the minimum acceptable performance to be the desired power.
```{r}
#set.seed(123)
result <- simulate_custom(
  data_function = my_data_generator,
  model_function = my_model_function,
  metric_function = metric_test_for_hte,
  minimum_acceptable_performance = 0.8, # 80% power
  mean_or_assurance = "mean",
  test_n = 100, # set to be low as performance metric doesn't use any test data.
  n_reps_total = 1000,
)

print(result)
plot(result)
```

### What if we are interested in a different performance metric?
So far we have calculated the minimum sample size for having adequate power to detect whether there are hetrogenous treatment effects using a statistical test. What if we are interested in other aspects of model performance?

I'll define a separate metric function and use this. For this we'll use the predieval package.

```{R}
metric_calib_slope_for_benifit <- function(test_data, fitted_model, model_name){
  x_names <- names(test_data)[grepl("^X", names(test_data))]
  test_data_trt1 <- test_data
  test_data_trt1$trt <- 1
  
  test_data_trt0 <- test_data
  test_data_trt0$trt <- 0
  
   capture_output(perf <- predieval::predieval(
    repeats = 5,
    Ngroups = 5, 
    X = test_data[, x_names, drop = FALSE],
    treat = test_data$trt,
    Y = test_data$Y,
    predicted.treat.1 = predict(fitted_model, newdata = test_data_trt1),
    predicted.treat.0 = predict(fitted_model, newdata = test_data_trt0),
    type = "continuous",
    bootstraps = 1 # set low as not interested in CI
  ))
  calib_slope_string <- perf$regression.for.benefit[[1]]
  calib_slope <- as.numeric(regmatches(calib_slope_string, regexpr("[0-9]+\\.?[0-9]*", calib_slope_string)))
  return(calib_slope)
}

```
### Large sample performance
Before running pmsims check what is possible in a large sample
```{r}

  set.seed(1234)
  large_sample_data <- my_data_generator(n = 50000)
  test_data <- my_data_generator(n = 5000) # set low as perf metric can be slow
  test_model <- my_model_function(large_sample_data)
  large_sample_performance <- metric_calib_slope_for_benifit(test_data, test_model, "multivariable regression")
  print(large_sample_performance)
```

### Running Simulate Custom for calibration slope
Now lets see what the minimum sample size to have a calibration slope for benifit of 0.9.
```{r}
#set.seed(123)
result <- simulate_custom(
  data_function = my_data_generator,
  model_function = my_model_function,
  metric_function = metric_calib_slope_for_benifit,
  minimum_acceptable_performance = 0.9, 
  mean_or_assurance = "assurance",
  test_n = 5000, # set low as perf metric can be slow
  n_reps_total = 200,
)

# what happens when sarting values fails to work

print(result)
plot(result)
```