---
title: "Simulate Custom"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulate Custom}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(pmsims)
```

## Motivation
The `simulate_custom()` function allows users to define their own data generation, model fitting, and performance metric  functions. This provides flexibility to calucate sample sizes for any prediction model, data type or performance metric. The `pmsims` built in sample size functions eg. `simulate_binary()` use simulate custrom under the hood.

## Usage
To use `simualte_custom()`, you need to provide three key functions:
1. **Data Function**: A function that generates synthetic datasets based on specified parameters.
2. **Model Function**: A function that fits a prediction model to the generated data.
3. **Metric Function**: A function that calculates the performance metric of interest.

In this example we will define a data function from a pre-existing dataset, use lasso regression, and assess performance using Brier Score.

### Defining the data generator
For this example we will use the built in mtcars data set. We will use the synthpop package to generate a very large synthetic dataset. We will then sample this dataset to give a required dataset of any size.

```{r}
set.seed(1234)
real_data = datasets::mtcars
synthetic_data <- synthpop::syn(real_data, k = 1000000, print.flag = FALSE)


my_data_generator <- function(n, data = synthetic_data$syn) {
  sample <-  data[sample(1:nrow(data), n, replace = FALSE), ]
  return(sample)
}

example_data <- my_data_generator(n = 10)
print(example_data)
```

### Defining the model function
Next we need to define the model function. The model function needs to take only the data returned by the data function as an argument. It needs to return a fitted model object that can be used with our metric function. we will use the glmnet package to fit an elastic net regression model, setting the elastic net mixing parameter to be 0.5. For this function the data must be in the form of a matrix. We look to predict mpg, which is the first column, with the remaining columns in the dataset. 

```{r}
my_model_function <-  function(data) {
      data <- as.matrix(data)
      
      data_matrix <- as.matrix(data)
      x <- data_matrix[, -1, drop = FALSE]
      y <- data_matrix[, 1]
      my_model <- glmnet::cv.glmnet(
        x,
        y,
        family = "gaussian",
        alpha = 0.5,
        nfolds = 10
      )
      return(my_model)
}
      
example_data <- my_data_generator(n = 100)
example__fitted_model <- my_model_function(example_data)
```


### Next we define a metric function. For this example we are going to use Mean Absolute Error. The metric function needs to take three positional arguments: test data, a fitted model, and a model name. The test data is used to evaluate the performance of the fitted model. The model name is a required argument even if unused. The metric function needs to return a single value representing the performance of the model on the test data. We return the negative of MAE as pmsims currently looks to optimise performance metrics where higher is better.

```{r}

my_metric = function(test_data, fitted_model, model_name) {
  test_data <- as.matrix(test_data)
  x_test <- test_data[, -1, drop = FALSE]
  y_test <- test_data[, 1]
  
  predictions <- predict(fitted_model, newx = x_test, s = "lambda.min")
  
  mae <- -mean(abs(predictions - y_test)) # set negative as pmsims expects higher performance to be better
  
  return(mae)
}

test_data = my_data_generator(n = 1000)
my_metric(test_data, example__fitted_model, "elastic net regression")
```

### What is the large sample performance
Before running pmsims it is important to understand the large sample performance of your model. When calcualting the minimum sample size we are looking for the smallest sample size at which the performance of the model meets a minimum acceptable performance. This minimum acceptable performance must be less than the performance that could be acheived with a large sample. The margin to the large sample performance will influence the minimum sample size returned. For this example we will use a large sample of 10000. For some machine learning models, in particular XGBoost, this may not be sufficient and larger samples should be used to evaluate 'large sample performance'.

```{r}

  set.seed(1234)
  large_sample_data <- my_data_generator(n = 10000)
  test_data <- my_data_generator(n = 10000)
  test_model <- my_model_function(large_sample_data)
  large_sample_performance <- my_metric(test_data, test_model, "elastic net regression")
  print(large_sample_performance)
```

We'll also understand small sample performance - this is what happens when we don't have a lot of data. 
```{r}
  set.seed(1234)
  small_sample_data <- my_data_generator(n = 50)
  test_data <- my_data_generator(n = 10000)
  test_model <- my_model_function(small_sample_data)
  small_sample_performance <- my_metric(test_data, test_model, "elastic net regression")
  print(small_sample_performance)
```
### Running Simulate Custom
We are now ready to run simulate custom. We decide that a mean absolute error of 2.6 is the minimum acceptable performance, allowing some degradation from the large sample performance. W For speed, will set the total number of replications to 500, although 1000 is recomended. This means that we will evaluate 20 different sample sizes in our search. We use default arguments for all other parameters. 

```{r}
#set.seed(123)
result <- simulate_custom(
  data_function = my_data_generator,
  model_function = my_model_function,
  metric_function = my_metric,
  minimum_acceptable_performance = -2.6,
  n_reps_total = 1000,
)

print(result)

```

