---
title: "Defining custom simulation workflows with `simulate_custom()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Defining custom simulation workflows with `simulate_custom()`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Motivation

The `simulate_custom()` function allows users to define their own data
generation, model fitting, and performance metric functions. This provides
flexibility to calculate sample sizes for any prediction model, data type or
performance metric. The `pmsims` built-in sample size functions---such as
`simulate_binary()`---use `simulate_custom()` under the hood.

For this vignette, we'll use five packages:

1. `pmsims`
2. `mlbench` package, which will provide the `PimaIndiansDiabetes` dataset
3. `synthpop` package, which will allow us to generate a large synthetic
   dataset based on the `PimaIndiansDiabetes` dataset.
4. `glmnet` for the elastic net modelling function.
5. `DescTools`, which provides the Brier score function used to assess model
   performance.

We will try to estimate the minimum sample size required for a prediction model
to predict `diabetes`.

```{r setup, message=FALSE}
library(pmsims)
library(mlbench)
library(synthpop)
library(DescTools)
library(glmnet)
```

## Using `simulate_custom()`

The function requires three key functions:

1. **Data function**: Generates synthetic datasets based on specified
   parameters.
2. **Model function**: Function that fits a prediction model to the generated
   data.
3. **Metric function**: Calculates the performance metric of interest.

In this example we will define a data function from a pre-existing dataset, use
elastic net regression, and assess performance using the Brier Score,
a performance metric for binary outcomes similar to mean squared error.

### Defining the data generator

For this example we will use the `PimaIndiansDiabetes` data set from the
`mlbench` package. The dataset consists of diabetes test results. For more
details, see the [help page][pima]. We will use the `synthpop` package to
generate a very large synthetic dataset. We will then sample from this dataset
to obtain a dataset of any required size.

[pima]: https://search.r-project.org/CRAN/refmans/mlbench/html/PimaIndiansDiabetes.html

```{r}
set.seed(1234)
data("PimaIndiansDiabetes", package = "mlbench")
real_data <- PimaIndiansDiabetes
real_data$diabetes <- ifelse(real_data$diabetes == "pos", 1, 0)

synthetic_data <- synthpop::syn(
  real_data,
  k = 1000000,
  print.flag = FALSE,
  minnumlevels = 2
)

my_data_generator <- function(n, data = synthetic_data$syn) {
  data[sample(seq_len(nrow(data)), n, replace = FALSE), ]
}

example_data <- my_data_generator(n = 10)
print(example_data)
```

### Defining the model function

Next, we need to define a model function. The model function needs to take
only the data returned by the data function as an argument. It must return
a fitted model object that can be used with our metric function.

We will use the `glmnet` package to fit an elastic net regression model, setting
the elastic net mixing parameter to 0.5. For this function, the data must be in
the form of a matrix. We aim to predict `diabetes` using the remaining columns
in the dataset.

```{r}
my_model_function <- function(data) {
  data_matrix <- as.matrix(data)
  outcome <- "diabetes"
  x <- data_matrix[, colnames(data_matrix) != outcome, drop = FALSE]
  y <- data_matrix[, outcome]

  glmnet::cv.glmnet(
    x,
    y,
    family = "binomial",
    alpha = 0.5,
    nfolds = 10
  )
}

example_data <- my_data_generator(n = 100)
example_fitted_model <- my_model_function(example_data)
fitted_model <- my_model_function(example_data)
```

### Metric function

Next, we define a metric function. For this example we use the Brier score.

The metric function must take three positional arguments:

`test_data`
: The data used to evaluate model performance.

`fitted_model`
: The fitted model object returned by the model function.

`model_name`
: The model name string. This argument is required even if unused.

The metric function must return a single numeric value representing model
performance on the test data. Here we return the negative Brier score, because
by default `pmsims` assumes that higher values indicate better performance.
If your custom metric may fail for some datasets, you can optionally set
`attr(my_metric, "value_on_error")` to define the fallback value returned for
failed simulation runs.

```{r}
my_metric <- function(test_data, fitted_model, model_name) {
  test_data_matrix <- as.matrix(test_data)
  y <- which(names(test_data) == "diabetes")
  x_test <- test_data_matrix[, -y]
  y_test <- test_data_matrix[, y]
  predictions <- predict(
    fitted_model,
    newx = x_test,
    s = "lambda.min",
    type = "response"
  )

  brier_score <- DescTools::BrierScore(y_test, pred = predictions)
  return(-brier_score)
}

test_data <- my_data_generator(n = 1000)
my_metric(test_data, example_fitted_model, "elastic net regression")
```
### What is the maximum achievable performance?

Before running `pmsims`, it is important to estimate the *maximum achievable
performance* of your modelling strategy. When using the wrapper functions (e.g.
`simulate_binary()`), this is provided directly by the user through arguments
such as `maximum_achievable_cstatistic`. In `simulate_custom()`, you must
estimate the maximum achievable performance yourself using a very large training
sample. You can then use this estimate to choose your `target_performance`.

When calculating the minimum sample size, we are looking for the smallest sample
size at which model performance meets a minimum acceptable threshold. This
threshold is passed through `target_performance` and should be set below the
maximum achievable performance. The gap between these two quantities will
influence the minimum sample size returned.

In this example, we use a training sample of 10,000 to approximate the maximum
achievable performance. For some machine learning models, particularly XGBoost,
this may be insufficient, and larger samples may be needed.

```{r}
set.seed(1234)
maximum_achievable_data <- my_data_generator(n = 10000)
test_data <- my_data_generator(n = 30000)
test_model <- my_model_function(maximum_achievable_data)
maximum_achievable_performance <- my_metric(
  test_data,
  test_model,
  "elastic net regression"
)
print(maximum_achievable_performance)
```

We will also look at small-sample performance, which reflects what happens when
we have limited data. We run this a few times because small-sample performance
can be highly variable.

```{r}
set.seed(1234)
small_sample_performance <- rep(NA, 10)
for(i in 1:10) {
  small_sample_data <- my_data_generator(n = 50)
  test_data <- my_data_generator(n = 30000)
  test_model <- my_model_function(small_sample_data)
  small_sample_performance[i] <- my_metric(
    test_data,
    test_model,
    "elastic net regression"
  )
}

print(small_sample_performance)
mean(small_sample_performance)
```

### Running `simulate_custom()`

We are now ready to run `simulate_custom()`. We decide that a Brier score of
-0.165 is the minimum acceptable performance, allowing some degradation from
the estimated maximum achievable performance. This target is approximately
halfway between `maximum_achievable_performance` and the small-sample
performance. For the Brier score there are no established criteria for choosing the
target, so it may be useful to explore how the estimated minimum sample size
changes across different values of `target_performance`. For speed, we will set
the total number of replications to 1000. We use default arguments for all
other parameters.

```{r, message=FALSE, warning=FALSE}
set.seed(1234)
result <- simulate_custom(
  data_function = my_data_generator,
  model_function = my_model_function,
  metric_function = my_metric,
  target_performance = -0.165,
  n_reps_total = 1000,
  progress = FALSE
)
print(result)
```

### Interpretation

The results show a minimum sample size of `r if (exists("result")) result$min_n else "not available in this render"`.
This is calculated using the assurance criterion, which means that we would
expect 80% of models developed on samples of this size to have a negative Brier
score of -0.165 or better.

Note that this is the minimum sample size required for the Brier score; other performance metrics may require larger sample sizes to achieve adequate performance.
